Extensible Gradual Typing Inside a Python

### 2014-10-12 atlang-modularity15/atlang-modularity15.tex
Renamed to 15

\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
%\usepackage{amsthm}
%\newtheorem{corollary}{Corollary}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
 \usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{pdflscape}
 \usepackage{colortbl}%
 \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
 \usepackage{wasysym}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, cls, @classmethod, with},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  numbersep=0.5em,
  xleftmargin=1.5em,
  stringstyle=\color{mauve},
  escapeinside={^}{^},
  aboveskip=1pt,
  belowskip=1pt
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`}
}
\lstdefinestyle{OpenCL}{
  language=C++,
  morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

@@

\title{Active Type Constructors}
\subtitle{Unambiguously Growing a Static Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers justifiably prefer abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests that a language is needed that is \emph{backwards compatible} with the libraries and tools of a widely adopted language, but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.

Our aim in this paper is to 1) introduce such a mechanism, and 2) show how it can itself be implemented as a library for an existing widely used language. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term constructor by delegating control, according to a novel bidirectional typechecking and translation protocol, to methods associated with a relevant \emph{active type constructor}, defined in a library. The delegate is stable even when additional extensions are imported, so we avoid ambiguities by construction. We demonstrate with several examples of interesting type system fragments as libraries.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when they require strengthening the language's static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in a foreign language requires programming against (and thus revealing) its internal implementation. This is brittle, unnatural and unsafe and can cause performance issues due to additional necessary checks and conversions at language boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues can prevent even dialects introducing only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This speaks to a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Internally extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics from within libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a distinct dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} (which permits syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of ambiguous extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the ambiguity.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

\paragraph{Terminology}
A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g. a fragment defining binary product types, organized around the type constructor $\fvar{prod}$). The syntax of each fragment can be characterized by its contributions to the \emph{concrete syntax} (e.g. expression forms \lstinlinep{(e, e)}, \lstinlinep{e.l} and \lstinlinep{e.r} and type form \lstinlinep{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g. $\FF{pair}[\FF{triv}](e_1; e_2)$ or $\FF{prj}[\FF{left}](e)$. Term indices are evaluated, if necessary, in an earlier \emph{phase} than term arguments and trivial indices are often omitted for clarity of presentation. The indices of expression terms,  like these examples,  are \emph{static terms}, $\sigma$. \emph{Types}, $\tau$, can be seen as static terms constructed by static term constructor $\FF{ty}$, indexed by a type constructor and having another (possibly trivial) static term as an argument, called (by convention) the \emph{type index}, e.g. $\FF{ty}[\fvar{prod}](\FF{spair}(\tau_1; \tau_2))$, often  abbreviated $\fvar{prod}\texttt{[}\tau_1; \tau_2\texttt{]}$ (%or stylized $\tau_1 \times \tau_2$. %Static terms are also called \emph{type-level terms}.  
 often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system. A type safe language is one where evaluation of a well-typed expression cannot lead to undefined behavior or result in a value of a different type, greatly simplifying static reasoning about program behavior and eliminating the need for many run-time checks.

%In a monolithic language design, decisions about which fragments are to be combined to constitute a complete  language are made centrally, so the designer can ensure that the concrete syntax is unambiguous and that other metatheoretic concerns, like guarantees of type safety, are maintained. To decentralize control over these decisions, so that fragments can be defined as libraries and combined arbitrarily, these guarantees must be shown to hold for all possible combinations of fragments deemed individually valid. This can quickly lead to difficulties.

  Each fragment also defines rules that contribute to the \emph{static semantics} (specifying how types are assigned to expressions at ``compile-time'') and the \emph{dynamic semantics} (specifying the behavior of well-typed expressions at ``run-time'') \cite{pfpl}. The dynamic semantics of full-scale languages are generally implemented (or even directly specified, as in the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}) by elaboration to a simpler \emph{internal language}.% In some languages, static terms themselves may have their own static and dynamic semantics (the \emph{static statics} and the \emph{static dynamics}).

The most apparently difficult problems related to combining fragments to form a typed programming language arise when attempting to enable the decentralized introduction of new concrete forms or term constructors. Syntactic ambiguities are difficult to modularly preclude and many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Fewer tools operate by exhaustive case analysis over type constructors. In a conventionally formulated programming language, however, it is difficult to introduce a new type system fragment without adding new concrete syntax or term constructors to accompany the new type constructors. In this paper, we aim to show that this need not be the case. We introduce a method that allows library providers to introduce new type system fragments by defining new type constructors in libraries while leaving the concrete syntax, the term constructors and the internal language fixed and enriching the static language so that it has its own dynamic semantics (the \emph{static dynamics}). The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a static function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label) corresponding to the now generalized abstract term  $\FF{attr}[\text{``r''}](e)$. Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'', $e$. As a result, tuple projection, as described above, can be introduced without the need to consider that it might syntactically coincide with projection out of a functional record or an object with a field labeled \lstinlinep{r} (both examples we will discuss). Each of these abstractions can use the same form and term constructor. As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional imports, so semantic ambiguities cannot arise by construction. %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

@@

\paragraph{Specific Aims} Our specific aims in this paper are to introduce active type constructors and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded. Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms. We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e. normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience). We describe two artifacts in pursuit of these aims:

@@

\item Because we are committed to a fixed syntax, but aim to show that our mechanism does not critically rely on any particular syntactic choices, it is pragmatic to use an existing syntax for which tooling already exists. In this paper, we choose Python's syntax. Moreover, we use the Python language itself as both the static language and the internal language. This makes it easy for us to implement our language, called @\texttt{lang}, itself as a library inside Python, using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a degenerate static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}. Taken together, @\texttt{lang} is almost completely backwards compatible with the Python ecosystem with remarkably little effort: we inherit a large number of  existing tools for working with Python files, Python's substantial packaging infrastructure and access to all existing Python libraries. We introduce @\texttt{lang} with practical examples of fragments as libraries in Sec. \ref{usage}.
\item In Sec. \ref{theory}, we describe active type constructors in their essential form, developing a minimal calculus called @$\lambda$. The syntax provides only variables, variable binding constructs, type ascription and two generic term constructors. Perhaps surprisingly, we show how a conventional syntax like Python's can be recovered by a purely syntactic desugaring to these term constructors. The semantics for @$\lambda$ take the form of a \emph{bidirectionally typed elaboration semantics} where the available type constructors are tracked by a \emph{fragment context}. We use a simple lambda calculus as both our static and internal language to emphasize the fundamental character of the approach.

@@

%
Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which strong correctness guarantees have been statically maintained using several type constructors defined in libraries. As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python, the static language. @\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy). This choice pays off immediately on the first line: @\texttt{lang}'s import mechanism is Python's import mechanism, so Python's build tools (e.g. \lstinlinep{pip}) and package repostories (e.g. \lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries, including those defining type constructors. Here, \lstinlinep{atlib} is the ``standard library'' but does not benefit from special support in \lstinlinep{atlang} itself.% Nothing in \lstinlinep{atlib} benefits from special support from \lstinlinep{atlang}, the core language package. We will see \lstinlinep{atlang} being used directly below. %@\texttt{lang} can be installed with a single command: \emph{omitted for review}. %@\texttt{lang} is agnostic about the particular design choices that we made in our examples.

%The top-level statements in an @\texttt{lang} file, like the \lstinlinep{print} statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. %That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of @\texttt{lang}. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
\subsection{Active Typechecking}

\paragraph{Types}
\begin{codelisting}[t]
\begin{lstlisting}
^\label{imports-start}^from atlib import record, decimal, dyn, string, 
  string_in, proto, fn, printf^\label{imports-end}^

@@

>>>>

\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
%\usepackage{amsthm}
%\newtheorem{corollary}{Corollary}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
 \usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{pdflscape}
 \usepackage{colortbl}%
 \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
 \usepackage{wasysym}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, cls, @classmethod, with},
  deletekeywords={min,max},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  numbersep=0.5em,
  xleftmargin=1.5em,
  stringstyle=\color{mauve},
  escapeinside={^}{^},
  aboveskip=1pt,
  belowskip=1pt
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`}
}
\lstdefinestyle{OpenCL}{
  language=C++,
  morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

@@

\title{Extensible Gradual Typing Inside a Python}
%\subtitle{Unambiguously Growing a Static Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests that a language is needed that is {backwards compatible} with the libraries and tools of a widely adopted language, but that has an  {internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported safely.

Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even when additional extensions are imported, so we avoid ambiguity by construction. We describe several examples of interesting type system fragments as libraries, including functional constructs, a type system for secure string sanitation and a foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several foreign languages often requires programming against a low-level internal implementation, rather than its semantics, via a foreign-function interface (FFI). This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries. As more languages are used together,  more FFIs must be developed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues can prevent even dialects introducing only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics and supporting the composition of these language fragments within or alongside libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

% \paragraph{Terminology}
% A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g. a fragment defining binary product types, organized around the type constructor $\fvar{prod}$). The syntax of each fragment can be characterized by its contributions to the \emph{concrete syntax} (e.g. expression forms \lstinlinep{(e, e)}, \lstinlinep{e.l} and \lstinlinep{e.r} and type form \lstinlinep{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g. $\FF{pair}[\FF{triv}](e_1; e_2)$ or $\FF{prj}[\FF{left}](e)$. Term indices are evaluated, if necessary, in an earlier \emph{phase} than term arguments and trivial indices are often omitted for clarity of presentation. The indices of expression terms,  like these examples,  are \emph{static terms}, $\sigma$. \emph{Types}, $\tau$, can be seen as static terms constructed by static term constructor $\FF{ty}$, indexed by a type constructor and having another (possibly trivial) static term as an argument, called (by convention) the \emph{type index}, e.g. $\FF{ty}[\fvar{prod}](\FF{spair}(\tau_1; \tau_2))$, often  abbreviated $\fvar{prod}\texttt{[}\tau_1; \tau_2\texttt{]}$ (%or stylized $\tau_1 \times \tau_2$. %Static terms are also called \emph{type-level terms}.  
%  often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system. A type safe language is one where evaluation of a well-typed expression cannot lead to undefined behavior or result in a value of a different type, greatly simplifying static reasoning about program behavior and eliminating the need for many run-time checks.

% %In a monolithic language design, decisions about which fragments are to be combined to constitute a complete  language are made centrally, so the designer can ensure that the concrete syntax is unambiguous and that other metatheoretic concerns, like guarantees of type safety, are maintained. To decentralize control over these decisions, so that fragments can be defined as libraries and combined arbitrarily, these guarantees must be shown to hold for all possible combinations of fragments deemed individually valid. This can quickly lead to difficulties.

%   Each fragment also defines rules that contribute to the \emph{static semantics} (specifying how types are assigned to expressions at ``compile-time'') and the \emph{dynamic semantics} (specifying the behavior of well-typed expressions at ``run-time'') \cite{pfpl}. The dynamic semantics of full-scale languages are generally implemented (or even directly specified, as in the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}) by elaboration to a simpler \emph{internal language}.% In some languages, static terms themselves may have their own static and dynamic semantics (the \emph{static statics} and the \emph{static dynamics}).

Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing, widely-used language. The most immediately difficult problems related to composing type system fragments arise when attempting to enable the decentralized introduction of new concrete syntax or corresponding term constructors to support the operations of this new type system. Syntactic ambiguities are difficult to modularly preclude (though some techniques exist). Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Fewer tools operate by exhaustive case analysis over type constructors. In a conventionally formulated programming language, however, it is difficult to introduce a new type system fragment without adding new concrete syntax or term constructors to accompany the new type constructors. In this paper, we aim to show that this need not be the case. We introduce a method that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete syntax, the term constructors and the target language of compilation fixed. The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a statically evaluated function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label). A fragment defining statically typed pairs might want this to refer to projection of the right component of the pair, while a fragment defining functional record types might want it project out the field labeled \verb|r| and an object system might introduce a more complex semantics involving dynamic dispatch. Were we to naively ``open up'' the syntax and semantics of the language, these fragments would conflict.

Instead, we assume that the syntax is fixed and includes this commonly used form, mapping it to abstract term  $\FF{attr}[\text{``r''}](e)$ (i.e. the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$). Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'' of the operation, $e$. Tuples, records, objects and variants thereof can all be supported over a common syntax. As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against (consistent with how extensible parsing is done in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction. %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

@@

\paragraph{Specific Aims} Our specific aims in this paper are to introduce active type constructors and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded. Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms. % We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e. normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience). 
We describe two artifacts in pursuit of these aims:

@@

\item We build our mechanism as a library inside Python, \verb|atlang|,  using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a  static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}. Thus, our mechanism represents a modularly extensible gradual type system. Taken together, @\texttt{lang} is almost completely backwards compatible with the Python ecosystem with remarkably little effort: we inherit a large number of  existing tools for working with Python files, Python's substantial packaging infrastructure and access to all existing Python libraries. We introduce @\texttt{lang}  in Sec. \ref{usage}. We continue in Sec. \ref{examples} by briefly describing  some more sophisticated examples, including statically typed functional datatypes with nested pattern matching and a complete and statically safe FFI to OpenCL for GPU programming.
\item In Sec. \ref{theory}, we describe active type constructors in their essential form, developing a minimal calculus called @$\lambda$. The syntax provides only variables, variable binding constructs, type ascription and two generic term constructors. Perhaps surprisingly, we show how a conventional syntax like Python's can be recovered by a purely syntactic desugaring to these term constructors. The semantics for @$\lambda$ take the form of a \emph{bidirectionally typed elaboration semantics} where the available type constructors are tracked by a \emph{tycon context}. We use a simple lambda calculus as both our static and internal language to emphasize the fundamental character of the approach. For more theoretically minded readers or those with no familiarity with Python, this section can be read first.

@@

\begin{codelisting}[t]
\begin{lstlisting}
^\label{imports-start}^from atlib import record, decimal, dyn, string, 
  string_in, proto, fn, printf^\label{imports-end}^

@@

\section{More Examples}\label{examples}
In the previous section, we showed examples of several interesting type system fragments implemented as libraries using \verb|atlang|, including functional record types, a prototypic object system and regular string types. A more detailed description of regular string types and their implementation using \verb|atlang| was recently published \cite{sanitation-psp14}. Here, we showcase two more powerful examples that demonstrate the expressive power of system: functional datatypes with nested pattern matching, and a type safe foreign function interface to the complete OpenCL language for many-core programming on devices like GPUs.
